Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-15 docs/manuscript.Rmd
updated

where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$

MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

@@

where $\beta_{MR} \neq 0$.

Measurement error of an exposure can be modeled as some transformation of the true value that leads to the observed value, $x_o = f(x)$. For example, we can define $f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}$, where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. Here the true value of the exposure is partially explained by the genetic instrument, $g$, such that

where $\beta_g$ is the effect of the SNP on the exposure, and $\epsilon_g$ is the normally distributed residual value of $x$; and the outcome is partially explained by the exposure

where $\beta_x$ is the true effect of the exposure on the outcome. In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that

When selecting an instrument $g$ that has a direct influence on $x$ where $x$ is causally related to $y$, one can reason that the influence of $g$ on $y$ is the proportion of the effect of $x$ on $y$ that is explained by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of $x$ on $y$ is estimated as

$$
\begin{aligned}
\beta_{MR} = \frac{\beta_{y_o}}{\beta_g} & = \frac{cov(y_o, g)}{cov(x_o, g)} \\
                                         & = \frac{\beta_{my} \beta_x cov(x, g)} {\beta_{mx} cov(x, g)} \\
                                         & = \frac{\beta_{my}} {\beta_{mx}} \beta_x
\end{aligned}
$$

Thus we obtain an estimate of the effect of $x_o$ on $y$ that is scaled by the measurement error in $x$ and $y$, but is unrelated to measurement imprecision. Measurement error and imprecision are likely to reduce the power of MR inferring a causal association (increased false negative rate), but importantly the evidence for causal association, based on rejecting the null hypothesis that $\beta_{MR} = 0$, is not inflated when the null hypothesis is true.

@@

>>>>

where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

@@

where $\beta_{MR} \neq 0$ denotes the existence of causality, and the $\beta_{MR}$ is an estimate of the causal effect.

Measurement error of an exposure can be modeled as a transformation of the true value that leads to the observed value, $x_o = f(x)$. For example, following Pierce and VanderWeele [-@Pierce2012] we can define

$$
f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}
$$

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. The same model of measurement error can be applied to the outcome variable $y$. In this study we assume there is no measurement error in the SNP.

@@

### Mediation-based causal inference under measurement error

In the causal inference test (CIT), an omnibus p-value is generated from four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is associated with $y$; 3) $x$ is associated with $y$; and 4) $g$ is independent of $y|x$. The 4th condition employs mediation for causal inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where $\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o$. When measurement error in scale and imprecision is introduced, such that $y_o$ is the measured value of $y$, it can be shown using basic covariance properties (Appendix 1) that
